Viscous Transport in Eroding Porous Media

### 2019-09-18 docs/transport/erosion_transport.tex
working on edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Flow in many natural occurring geometries involves flows that vary over
many orders of magnitude.  \todo[]{The grammar in the first sentence is awkward: Flow in ... involves flows?}
For example, the length scales of a porous
media can vary from $10^{-6}m$ to
$10^{-1}m$~\cite{mil-chr-imh-mcb-ped1998}.  Porous media flow plays an
important role in several environmental and industrial applications.
Therefore, understanding the dynamics of pore-scale flow is central to
upscale the governing fluid equations, characterize
dispersion~\cite{saf1959}, and quantify mixing. However, because of the
large disparity of length and velocity scales, constitutive laws that
link the microscopic scale to macroscopic scale must be
developed~\cite{mil-chr-imh-mcb-ped1998}.  Therefore, to find closure
formulas, accurate simulations at microscopic scales must be performed
and analyzed.  A few examples of closure formulas and coarse-grained
variables include
pressure-saturation-permeability~\cite{mil-chr-imh-mcb-ped1998},
permeability-porosity~\cite{dar-mcc1998, car1937, won-kop-tom1984},
first passage time distribution~\cite{ber-sch-sil2000,
hym-den-hag-kan2019, cve-che-wen1996}, tortuosity~\cite{hak-com-den2019,
mat-kha-koz2008, dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, and anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}.

In many applications, the fluid transport continuously changes the pore
structure.  This is particularly common in geophysical flows where
erosion alters the porous matrix.  In this work, we investigate flow in
eroded porous media.  By introducing erosion, there is a two-way
coupling between the hydrodynamics and the mechanical alteration of the
complex porous geometry.  Transport in porous geometries that have
undergone erosion finds applications in aquifers~\cite{cve-che-wen1996,
sud1986}, solute transport in groundwater flow~\cite{dag1987,
kon-bre1978}.  Porous geometries that have undergone erosion have
certain characteristics.  For example, the geometry often has regions of
high porosity that connect the inlet to the outlet.  This channelization
results in velocity scales that can vary over several orders of
magnitude~\cite{all-hea-lab-rei2002}.  Moreover, the channelization
results in a geometry that is largely heterogeneous and anisotropic, and
this affects heat transfer through the medium~\cite{nil-sto1990,
ree-sto1995}.  Others have studied transport in heterogeneous geometries
(for example, see~\cite{mil-chr-imh-mcb-ped1998, ber-sch-sil2000,
hak-com-den2019, suo-liu-gan2019, cus-hu-den1995, cve-che-wen1996,
leb-ded-dav-bou2007, knu-car2005, bel-sal-rin1992}). However, to the
best of our knowledge, a detailed study of transport in eroded
geometries has not been performed.

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
hydrodynamics, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018, par-izu2000}.  A variety of
numerical methods can be applied to solve the incompressible Stokes
equations such as a lattice-gas or lattice Boltzmann
model~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volume~\cite{suo-liu-gan2019, den-ica-hid2018, sie-ili-pri-riv-gua2019},
finite differences~\cite{leb-ded-dav-bou2007, knu-car2005}, and cellular
automata~\cite{rot1988}. In this work, we instead solve an equivalent
boundary integral equation (BIE) formulation of the incompressible
Stokes equation.  The BIE is discretized with a high-order quadrature
method that resolves nearly touching bodies with a near-singular
integration scheme~\cite{bar-wu-vee2015}.  The erosion dynamics are
simulated by applying a stable second-order time stepping Runge-Kutta
method to the {\thL} formulation~\cite{hou-low-she1994} of the eroding
bodies.  Given the solution of the incompressible Stokes equations are
solved in an eroded porous geometry, we investigate different metrics to
characterize transport.  We focus on three metrics: the tortuosity, the
anomalous dispersion, and the distribution of geometrical channels.

The first two of these metrics can be defined in terms of statistical
moments of trajectories.  Depending on the application, transport at the
microscale can be modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019}, and advection
dispersion equation~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion).  Therefore, a trajectory $\ss(t)$
initialized at $\ss_0$ is governed by
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no diffusion, all
dispersion is a result of the particle spreading induced by the porous
geometry.  Since there is no diffusion, we apply a high-order explicit
Runge-Kutta time stepping method to compute streamlines.  However, other
groups have simulated trajectories using continuous time random
walks~\cite{mon-wei1965, den-cor-sch-ber2004, leb-den-car2008,
ber-cor-den-sch2006} and a Fokker-Planck equation~\cite{ica-den2019}.

\todo[inline]{Dump of text}
Resolving the flow in these small velocity regions is essential to
accurately describe anomalous dispersion and
mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}.
\todo[inline]{Dump of text}

In a porous geometry, the length of a particle path is larger than the
distance travelled if the grains were absent.  To quantify the amount of
winding a particle takes, we use the tortuosity that is defined as the
ratio of the particle paths length to the distance between the inlet and
outlet.  In fractured media, the local tortuosity can be over
2.5~\cite{hym-den-hag-kan2019}, and in porous media, depending on the
porosity, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, mat-kha-koz2008}, or greater than
2~\cite{dud-koz-mat2011}.  The tortuosity of the geometry is defined by
averaging over the entire inlet.  Therefore, the geometry's tortuosity
can be used to characterize average particle
motions~\cite{hak-com-den2019}.

The tortuosity is a first statistical moment of the streamlines.
However, to measure spread of trajectories, second-order moments are
computed. By computing the variance of the lengths of trajectories in a
porous media, non-local anomalous dispersion is
observed~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}, resulting in super-dispersion.  The
dispersion depends on the amount of time trajectories spend in low and
high velocity regimes~\cite{ber-sch2001}.  Anomalous dispersion has been
investigated for a variety of porous geometries, but not in the context
of eroding porous geometries.

Rather than solving the fluid equations in a complex porous matrix,
network models can be used to describe the flow.  Here, the pores
between neighboring grains are thought of as small channels, and the
geometry consists of a connected network of these
channels~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006} with the
flow resembling a Hagen-Poiseuille profile in each channel.  So that
network models can be applied to an eroding geometry, we characterize
the pore sizes between neighboring grains as a function of the
geometry's porosity.  The pore size distribution can be used to quantify
velocity distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\cite{sie-ili-pri-riv-gua2019}, and
connectivity~\cite{knu-car2005, wes-blo-gra2001}.

The numerical methods to solve the governing fluid and erosion equations
are based on previous work of the authors~\cite{qua-moo2018}.  We
continue to use a boundary integral equation (BIE) formulation of the
incompressible Stokes equations and a second-order Runge-Kutta method
applied to a smoothed version of the magnitude of the shear stress.  In
our previous work, we discretized the BIE with the trapezoid rule whose
accuracy grows when evaluating layer potentials close to the fluid
boundary.  Therefore, given a fixed resolution, our previous work
required a minimum separation distance between all bodies to maintain
stability.  One of the earliest works on near-singular integration was
developed by Baker and Shelley~\cite{bak-she1986}, and in recent years,
many other near-singular integration schemes have been
developed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  In this work, we use a Barycentric
quadrature method~\cite{bar2014, bar-wu-vee2015} since it is a
non-intrusive modification of the trapezoid rule, and the error is
guaranteed to be uniformly bounded.  In this work, we extend this
quadrature method so that it can be used to compute the gradient of the
velocity which is required to compute the shear stress and the fluid
vorticity. 
\begin{figure}[H]
\begin{center}
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b1}
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b80}
\includegraphics[width = 0.38 \textwidth]{./figs/vort_50b160}\\
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b240}
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b320}
\includegraphics[width = 0.38 \textwidth]{./figs/vort_50b400}
\caption{\label{fig:Eroding50vort} Erosion of 50 nearly touching
bodies with the fixed pressure drop condition. The color is the
vorticity of the fluid. The 6 snapshots are evenly spaced in time.}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
Our first main contribution is extending the work of Barnett et
al.~\cite{bar-wu-vee2015} so that high-order accurate derivatives of
solutions of the Stokes equations can be computed.  Using this new
quadrature method, we can accurately compute erosion rates, and
therefore stably simulate erosion.  Moreover, we use the quadrature
method to visualize the vorticity in the fluid bulk.  The vorticity is
convenient to visualize since, in the zero Reynolds number limit, it is
equivalent to the shear stress on the eroding bodies, and therefore its
magnitude is the erosion rate.

Our second contribution is computing trajectories in the eroded
geometries.  These trajectories are used to characterize the transport
by computing local tortuosities of the trajectories, the tortuosity of
the eroded geometry, and the anomalous dispersion rates in the geometry.
Again, by using an appropriate quadrature method~\cite{bar-wu-vee2015},
and a fourth-order Runge-Kutta method, we are able to stably compute
trajectories that spend large amounts of time in the slow velocity
regions near the eroding bodies, and computing these trajectories
reliably is essential to characterize the flow.  We compare our
tortuosity and dispersion rates with models available in the literature.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}
As in our previous work~\cite{qua-moo2018}, this work considers
two-dimensional immobile bodies. However, this new work does not have a
limitation on the packing of the bodies. Compared to our previous work,
we are able to simulate much denser suspensions with a comparable number
of discretization points.  While we do investigate transport in this
work, we only consider passive particles that are undergoing only
advection. To model other physical processes, such as contaminant
transport, we plan to simulate a finite Peclet number
advection-diffusion equation in eroded geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we briefly summarize the model that we
describe in detail in our previous work~\cite{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:method} describes the numerical methods, with special
attention paid to the new quadrature method to compute the shear stress
and vorticity.  Section~\ref{sec:transport} describes measures for
characterizing the geometry, transport, and dispersion.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  A more
detailed description is in our previous work~\cite{qua-moo2018}.  We
consider flows inside a confined geometry $\Omega$ that contains $M$
eroding bodies with boundary $\gamma_\ell$, $\ell = 1,\ldots,M$.  The
boundary of the bounding geometry is $\Gamma$, and the fluid domain is
bounded by $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$.  On each eroding body, $\nn$ is the normal vector pointing
into the body, and $\ss$ is the unit tangent vector pointing in the
counterclockwise direction.  The normal vector on $\Gamma$ is the
outward unit normal.  Neglecting inertial forces, the governing
equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}\\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass} \\
    \uu = 0, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the bodies} \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity} \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$.

The erosion rate loses differentiability if the shear stress changes
sign, and this leads to corners developing on $\gamma$ and numerical
instabilities.  Therefore, we modify the erosion rate by adding a
curvature penalization. That is, the erosion model
in~\eqref{eqn:erosionModel} is replaced with 
\begin{align}
  \Vn = \CE \, \abs{\tau} + \epsilon \langle\abs{\tau}\rangle \left(
    \frac{L}{2\pi} \kappa - 1 \right),
\end{align}
where $\epsilon \ll 1,$ $\langle \cdot \rangle$ is the spatial average,
$L$ is the length of $\gamma$, and $\kappa$ is the curvature of
$\gamma$.  The normalization term guarantees that the regularization
only changes the body's shape, but not its total length.  Finally, to
increase the overall stability of the method, a narrow Gaussian filter
is applied to the erosion rate.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\cite{mil-chr-imh-mcb-ped1998}, velocity scales can
be as small as $10^{-1}\mathrm{m/day}$~\cite{kut-scr-dav-ham1995}, and
the scales of a single porous media can range over several orders of
magnitude. Numerical methods that accurately address all of these scales
are essential to develop meaningful constitutive relationships between
the microscopic and macroscopic scales~\cite{mil-chr-imh-mcb-ped1998},
to characterize dispersion~\cite{saf1959}, and to quantify
mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}.
Flow in porous media is further complicated when there is a two-way
coupling between the flow and the geometry. For example, the grains of a
porous media evolve in applications involving
melting~\cite{bec-vis1998}, dissolution~\cite{kan-zha-che-he2002},
deposition~\cite{joh-eli1995}, and biofilm
growth~\cite{tan-val-wer2015}.  We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical and hydrological
applications.

Examples of coarse-grained models for porous media flow include
pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.  Some of these works consider
heterogeneous porous geometries, but to the best of our knowledge, a
detailed study of transport in eroded geometries has not been performed.
This paper characterizes transport in eroded geometries by computing and
analyzing a variety of coarse-grained variables.

When a porous geometry erodes, certain qualitative characteristics are
unveiled that affect transport through the geometry. For example, an
eroded geometry often contains channels of high porosity that connect
the inlet to the outlet, and this results in velocities that vary over
several orders of magnitude~\cite{all-hea-lab-rei2002}.  Moreover,
channelized geometries are heterogeneous and anisotropic, and this
affects the transfer of quantities such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.  To characterize the transport in
eroded geometries, we must complete three main tasks.  The first two are
forming high-fidelity numerical solutions of the fluid equations, and
eroding the grains according to an erosion model. These two tasks are
performed using our previous work~\cite{qua-moo2018}, but we improve our
numerical methods so that simulations of more realistic dense
suspensions of eroding grains are possible
(Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  Finally,
in this paper, we simulate and characterize transport through eroded
porous geometries.

\begin{figure}[H]
\begin{center}
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b1}
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b80}
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b160}\\
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b240}
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b320}
\includegraphics[height = 0.31 \textwidth]{./figs/vel_log_50b400}
\caption{\label{fig:Eroding50vel} Six evenly spaced snapshots of 50
eroding bodies.  The background velocity is a Hagen-Poiseuille flow with
a varying flow strength that corresponds to a constant pressure drop
across the channel. The color is the magnitude of the velocity in a
logarithmic scale. The flow velocity varies over several orders of
magnitude, the grains form irregular shapes with large aspect ratios,
and the geometry becomes channelized and anisotropic.}
\end{center}
\end{figure}

Because of the length and velocity scales in hydrological applications
such as groundwater flow~\cite{bea1972}, we model the hydrodynamics with
the two-dimensional incompressible Stokes equations. A variety of
numerical methods are available to solve the incompressible Stokes
equations including lattice-gas or lattice
Boltzmann~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volumes~\cite{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\cite{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\cite{rot1988}.  However, to
accurately simulate erosion and transport, we require a high-order
method that is able to resolve geometries with arbitrarily shaped and
nearly-touching grains.  Therefore, we solve the incompressible Stokes
equations with a boundary integral equation (BIE).  To couple erosion
with the fluid, the grains are eroded at a rate that is proportional to
the magnitude of the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, par-izu2000}.  Since the vorticity and the
shear stress are equivalent on the grains in the zero Reynolds number
limit, we also compute the vorticity in the fluid bulk
(Figure~\ref{fig:Eroding50vort}).

To characterize the transport, particle trajectories need to be
computed. Depending on the application, microscale transport can be
modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019},
advection-diffusion~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion).  Therefore, a trajectory $\ss(t)$
initialized at $\ss_0$ is governed by
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity, and the trajectories are simply the
streamlines. Since there is no stiff diffusive term
in~\eqref{eqn:tracers}, we use a high-order explicit Runge-Kutta time
stepping method. Other methods to compute streamlines include a
continuous time random walk~\cite{mon-wei1965, den-cor-sch-ber2004,
leb-den-car2008, ber-cor-den-sch2006}.

\begin{figure}[H]
\begin{center}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b1}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b80}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b160}\\
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b240}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b320}
\includegraphics[height = 0.3 \textwidth]{./figs/vort_50b400}
\caption{\label{fig:Eroding50vort} The same six snapshots from
Figure~\ref{fig:Eroding50vel}.  The color is the vorticity of the fluid.
Since the rate of erosion is equivalent to the magnitude of the
vorticity, erosion is fastest in the yellow and blue regions and slowest
in the green regions.}
\end{center}
\end{figure}

To be able to compute stable simulations of erosion and accurate
streamline trajectories, we use high-order methods in both space and
time. The erosion rate is singular at stagnation points, so we apply a
mild regularization and smoothing terms to eliminate numerical
instabilities.  As described in our previous work~\cite{qua-moo2018}, we
apply a stable second-order Runge-Kutta method is applied to the {\thL}
coordinates~\cite{hou-low-she1994} of the eroding grains.  In these
coordinates, the stiff term is linear and treated implicitly, so we are
able to take large time step sizes. In our previous work, after
discretizing the BIE, the trapezoid rule is used to approximate the
integrals. The accuracy of the trapezoid rule is sufficiently accurate
if all the bodies are sufficiently separated~\cite{tre-wei2014}, but
nearly-singular integrals must be computed when two grains are
nearly-touching or when evaluating the velocity or vorticity near an
eroding grain.  One of the earliest works on near-singular integration
was developed by Baker and Shelley~\cite{bak-she1986}, and in recent
years, many other near-singular integration schemes have been
developed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  In this work, we use a Barycentric
quadrature method~\cite{bar2014, bar-wu-vee2015} since it is a
non-intrusive modification of the trapezoid rule, and the error is
guaranteed to be uniformly bounded.  We extend the original quadrature
method so that it can be used to compute the gradient of the velocity
that is required to compute the shear stress and the fluid vorticity. In
addition, the quadrature method is used to accurately compute streamline
trajectories that pass through the very slow regions close to the
eroding grains.

Once streamlines in an eroded geometry are computed, we characterize the
transport by computing and analyzing three different metrics: the
tortuosity, the anomalous dispersion, and the pore size distribution.
The local tortuosity of a streamline connecting the inlet to the outlet
is the streamline's length normalized by the distance between the inlet
and the outlet.  In porous media, depending on several factors including
the porosity, the local tortuosity can be greater than
1.5~\cite{kop-kat-tim1996, mat-kha-koz2008}, or even
2~\cite{dud-koz-mat2011}.  The tortuosity of a geometry is defined by
averaging the local tortuosity over all streamlines that are initialized
at the inlet, and the geometry's tortuosity can be used to characterize
average particle motions~\cite{hak-com-den2019}.  To characterize
spreading, the fluid dispersion is computed as the variance of the
streamline lengths. In porous media, this spreading can often be
characterized with a super-diffusive anomalous dispersion
rate~\cite{kan-dea-nun-bij-blu-jua2014, cus-hu-den1995,
dea-leb-den-tar-bol-dav2013}. An accurate solution near the grains is
crucial since the anomalous dispersion is caused by streamlines spending
time in both the high and low velocity regimes~\cite{ber-sch2001}.
Finally, the third metric, pore size distribution, is useful to quantify
velocity distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\cite{sie-ili-pri-riv-gua2019},
connectivity~\cite{knu-car2005, wes-blo-gra2001}, and to develop network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions}
Our first contribution is extending the work of Barnett et
al.~\cite{bar-wu-vee2015} so that high-order accurate derivatives of
solutions of the Stokes equations can be computed.  Using this new
quadrature method, we accurately compute erosion rates and stably
simulate erosion.  Moreover, we use the quadrature method to visualize
the vorticity in the fluid bulk.  The vorticity is convenient to
visualize since, in the zero Reynolds number limit, it is equivalent to
the shear stress on the eroding bodies, and therefore its magnitude is
the erosion rate.

Our second contribution is computing and analyzing trajectories in the
eroded geometries.  These trajectories are used to characterize the
transport by computing local tortuosities of the trajectories, the
tortuosity of the eroded geometry, and the anomalous dispersion rates in
the geometry.  By using the same quadrature rule and a fourth-order
Runge-Kutta method, we are able to accurately compute trajectories that
spend large amounts of time in the slow velocity regions near the
eroding bodies, and computing these trajectories reliably is essential
to characterize the flow~\cite{dea-leb-den-tar-bol-dav2013}.  We compare
our tortuosity and dispersion rates with models available in the
literature.  Finally, we analyze the effect of erosion on the pore size
distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations}
As in our previous work~\cite{qua-moo2018}, we consider two-dimensional
immobile eroding grains. However, this new work does not have a limitation on
the proximity of the bodies. Compared to our previous work, we are able to
simulate much denser suspensions with a comparable number of
discretization points and without upsampling.  While we do investigate
transport in this work, we only consider passive particles that are only
advected by the flow.  To model other physical processes, such as
contaminant transport, we plan to simulate a finite Peclet number
advection-diffusion equation in eroded geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the Paper}
In Section~\ref{sec:formulation}, we summarize the model that we
describe in detail in our previous work~\cite{qua-moo2018}.  In
Section~\ref{sec:DLP}, we recast all the governing equations as layer
potentials defined in both $\RR^2$ and in $\CC$.
Section~\ref{sec:transport} describes measures for characterizing the
geometry, transport, and dispersion. Section~\ref{sec:method} describes
the numerical methods, with special attention paid to the new quadrature
method to compute the shear stress and vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
Section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
summarize the model, and a more detailed description is in our previous
work~\cite{qua-moo2018}.  We consider flows inside a confined geometry
$\Omega$ that contains $M$ eroding bodies with boundaries $\gamma_\ell$,
$\ell = 1,\ldots,M$.  The boundary of the bounding geometry is $\Gamma$,
so the fluid domain is bounded by $\bd \Omega = \Gamma \cup \gamma_1
\cup \cdots \cup \gamma_M$.  In all examples, we let $\Gamma$ be a
slightly rounded-off version of $[-3,3] \times [-1,1]$, and we only
place eroding bodies in the region $[-1,1] \times [-1,1]$.  On each
eroding body, $\nn$ is the normal vector pointing into the body, and
$\ss$ is the unit tangent vector pointing in the counterclockwise
direction.  The normal vector on $\Gamma$ is the outward unit normal.
Neglecting inertial forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}\\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass} \\
    \uu = 0, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the bodies} \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity} \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here, $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, $\Vn$ is the normal velocity
of $\gamma$, and
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
is the shear stress on $\gamma$.  The erosion process is simulated by
alternating between solving the fluid equations and adancing the grains
according to the erosion rate for a small time step.

The erosion rate loses differentiability if the shear stress changes
sign, and this leads to corners developing on $\gamma$ and numerical
instabilities.  Therefore, we modify the erosion rate, $V_\nn$ in
equation~\eqref{eqn:erosionModel}, with
\begin{align}
  \Vn = \CE \, \abs{\tau} + \epsilon \langle\abs{\tau}\rangle \left(
    \frac{L}{2\pi} \kappa - 1 \right),
\end{align}
where $\epsilon \ll 1,$ $\langle \cdot \rangle$ is the spatial average,
$L$ is the length of $\gamma$, and $\kappa$ is the curvature of
$\gamma$.  This new term penalizations regions of high curvature, and
the normalization term guarantees that the regularization only changes
the body's shape, but not its total length.  Finally, to increase the
overall stability of the method, a narrow Gaussian filter is applied to
the erosion rate at each time step.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Boundary Integral Equation Formulation}
\label{sec:DLP}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Since require a method to solve the incompressible Stokes equations in
complex two-dimensional porous geometries, we reformulated the governing
PDE as a boundary integral equation. This has the advantage that only
the one-dimensional boundary of the domain must be discretized, and if
appropriate quadrature formula and fast summation methods are used, the
result is a high-fidelity numerical simulation that is computed with
near-optimal complexity.
